\pkg{partykit}: A Toolkit for Recursive Partytioning

### 2010-04-23 pkg/inst/doc/introduction.Rnw
splits more formally
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@167 255f4249-2263-4417-b1ea-8498779e27a7
cosmetics
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@168 255f4249-2263-4417-b1ea-8498779e27a7

\author{Torsten Hothorn\\Ludwig-Maximilians-Universit\"at M\"unchen
   \And Achim Zeileis\\Leopold-Franzens-Universität Innsbruck}
\Plainauthor{Torsten Hothorn, Achim Zeileis}

@@

A split is basically a function that maps data, 
more specifically a partitioning variable, to child nodes. 
Objects of class \code{partysplit} are designed to represent 
such functions and are set-up by the \code{partysplit()} 
constructor:
<<partysplit-1, echo = TRUE>>=
## binary split in numeric variable `Sepal.Length'
sl5 <- partysplit(which(names(iris) == "Sepal.Length"),
    breaks = 5)
class(sl5)
@
The internal structure of class \code{partysplit} contains
information about the partitioning variable, the split-points,
the handling of split-points, the treatment of observations with
missing values and the daughter nodes to send observations to:
<<partysplit-2, echo = TRUE>>=
unclass(sl5)
@
Here, the split is defined in the first variable (corresponds to
\code{Sepal.Length} in data frame \code{iris}) and the splitting rule
is \code{Sepal.Length} $\le 5$:
<<partysplit-3, echo = TRUE>>=
character_split(sl5, data = iris)
@
This representation of splits is completely abstract and, most importantly,
independent of any data. Now, data comes into play when we actually
want to perform splits:
<<partysplit-4, echo = TRUE>>= 
kidids_split(sl5, data = iris)
@
For each observation in \code{iris} the split is performed and the 
number of the daughter node to send this observation to
is returned. Of course, this is a very complicated way of saying
<<partysplit-5, echo = TRUE>>=
(!with(iris, Sepal.Length <= 5)) + 1
@

However, splits may become more complex. Consider a split in a categorical
variable at three levels where the first two levels go to the left
daugther node and the third one to the right daugther node:
<<partysplit-6, echo = TRUE>>=
## binary split in factor `Species'
sp <- partysplit(which(names(iris) == "Species"),
    index = c(1L, 1L, 2L))
character_split(sp, data = iris)
table(kidids_split(sp, data = iris), iris$Species)
@
The internal structure of this object now contains an index slot
<<partysplit-6, echo = TRUE>>=
unclass(sp)
@
that maps levels to daughter nodes. This mapping is also
useful with splits in ordered variables, for example when 
representing multiway splits:
<<partysplit-7, echo = TRUE>>=
## multiway split in numeric variable `Sepal.Width',    
## higher values go to the first kid, smallest values
## to the last kid
sw23 <- partysplit(which(names(iris) == "Sepal.Width"),       
    breaks = c(3, 3.5), index = 3:1)    
character_split(sw23, data = iris)
table(kidids_split(sw23, data = iris),
    cut(iris$Sepal.Width, breaks = c(-Inf, 2, 3, Inf)))
@
The mapping of classes of the categorized numeric variable
to daugther nodes can be changed by modifying \code{index}:
<<partysplit-8, echo = TRUE>>=
sw23 <- partysplit(which(names(iris) == "Sepal.Width"),
    breaks = c(3, 3.5), index = c(1L, 3L, 2L))
character_split(sw23, data = iris)
@

@@

The function $g$ essentially determines, which of the intervals (defined by $\mu$) the value
$x_i$ is contained in ($I$ denotes the indicator function here):
\begin{eqnarray*}
x \mapsto g(x) = \sum_{j = 1}^l j I_{A(j)}(x_i)
\end{eqnarray*}
where $A(j) = (\mu_{j - 1}, \mu_j]$ for \code{right = TRUE} except $A(l) = (\mu_{l - 1}, \infty)$.
If \code{right = FALSE}, then $A(j) = [\mu_{j - 1}, \mu_j)$ except 
$A(1) = (-\infty, \mu_1)$. Note that with some categorical variable $x_i$ and default
split points, $g$ is the identity.

In the simplest case of a binary split in a numeric variable $x_i$, there is only
one split point $\mu_1$ and, with $\sigma = (1, 2)$, observations with $x_i \le \mu_1$ are sent
to daugther node $d_1$ and observations with $x_i > \mu_1$ to $d_2$.

>>>>

\author{Torsten Hothorn\\Ludwig-Maximilians-Universit\"at \\ M\"unchen
   \And Achim Zeileis\\Leopold-Franzens-Universit\"at \\ Innsbruck}
\Plainauthor{Torsten Hothorn, Achim Zeileis}

@@

A split is basically a function that maps data, 
more specifically a partitioning variable, to daugther nodes. 
Objects of class \code{partysplit} are designed to represent 
such functions and are set-up by the \code{partysplit()} 
constructor:
<<partysplit-1, echo = TRUE>>=
## binary split in numeric variable `Sepal.Length'
sl5 <- partysplit(which(names(iris) == "Sepal.Length"),
    breaks = 5)
class(sl5)
@
The internal structure of class \code{partysplit} contains
information about the partitioning variable, the split-points,
the handling of split-points, the treatment of observations with
missing values and the daughter nodes to send observations to:
<<partysplit-2, echo = TRUE>>=
unclass(sl5)
@
Here, the split is defined in the first variable (corresponds to
\code{Sepal.Length} in data frame \code{iris}) and the splitting rule
is \code{Sepal.Length} $\le 5$:
<<partysplit-3, echo = TRUE>>=
character_split(sl5, data = iris)
@
This representation of splits is completely abstract and, most importantly,
independent of any data. Now, data comes into play when we actually
want to perform splits:
<<partysplit-4, echo = TRUE>>= 
kidids_split(sl5, data = iris)
@
For each observation in \code{iris} the split is performed and the 
number of the daughter node to send this observation to
is returned. Of course, this is a very complicated way of saying
<<partysplit-5, echo = TRUE>>=
(!with(iris, Sepal.Length <= 5)) + 1
@

The function $g$ essentially determines, which of the intervals (defined by $\mu$) the value
$x_i$ is contained in ($I$ denotes the indicator function here):
\begin{eqnarray*}
x \mapsto g(x) = \sum_{j = 1}^l j I_{\mathcal{A}(j)}(x_i)
\end{eqnarray*}
where $\mathcal{A}(j) = (\mu_{j - 1}, \mu_j]$ for \code{right = TRUE} except 
$\mathcal{A}(l) = (\mu_{l - 1}, \infty)$.
If \code{right = FALSE}, then $\mathcal{A}(j) = [\mu_{j - 1}, \mu_j)$ except 
$\mathcal{A}(1) = (-\infty, \mu_1)$. Note that with some categorical variable $x_i$ and default
split points, $g$ is the identity.

In the simplest case of a binary split in a numeric variable $x_i$, there is only
one split point $\mu_1$ and, with $\sigma = (1, 2)$, observations with $x_i \le \mu_1$ are sent
to daugther node $d_1$ and observations with $x_i > \mu_1$ to $d_2$. However,
this representation of splits is general enough to deal with more complicated
set-ups like surrogate splits, where typically the index needs modification, for example
$\sigma = (2, 1)$, categorical splits, i.e., there is one data structure for both
ordered and unordered splits, multiway splits, and functional splits. The latter
can be implemented by defining a new artificial splitting variable $x_{p + 1}$ by
means of a potentially very complex function of $x$ later used for splitting.

As an example, consider a split in a categorical
variable at three levels where the first two levels go to the left
daugther node and the third one to the right daugther node:
<<partysplit-6, echo = TRUE>>=
## binary split in factor `Species'
sp <- partysplit(which(names(iris) == "Species"),
    index = c(1L, 1L, 2L))
character_split(sp, data = iris)
table(kidids_split(sp, data = iris), iris$Species)
@
The internal structure of this object contains the \code{index} slot
<<partysplit-6, echo = TRUE>>=
unclass(sp)
@
that maps levels to daughter nodes. This mapping is also
useful with splits in ordered variables, for example when 
representing multiway splits:
<<partysplit-7, echo = TRUE>>=
## multiway split in numeric variable `Sepal.Width',    
## higher values go to the first kid, smallest values
## to the last kid
sw23 <- partysplit(which(names(iris) == "Sepal.Width"),       
    breaks = c(3, 3.5), index = 3:1)    
character_split(sw23, data = iris)
table(kidids_split(sw23, data = iris),
    cut(iris$Sepal.Width, breaks = c(-Inf, 2, 3, Inf)))
@
The mapping of classes of the categorized numeric variable
to daugther nodes can be changed by modifying \code{index}:
<<partysplit-8, echo = TRUE>>=
sw23 <- partysplit(which(names(iris) == "Sepal.Width"),
    breaks = c(3, 3.5), index = c(1L, 3L, 2L))
character_split(sw23, data = iris)
@

@@
